---
title: "Data visualisatin"
author: "Ahsan"
date: "2023-06-26"
output: html_document
---

```{r setup, include=FALSE}
library("grDevices")
library("tidyverse")
library("viridis")
library("showtext")
library("here")
library("ggpubr")
library("extrafont")
library("kableExtra")
library("tidyr")
library("lubridate")
library("readxl")
library("ggstatsplot")
library("pairwiseComparisons")
library("multcompView")
library("ggthemes")
library("emmeans")
library("multcomp")
library("car")
library("DHARMa")
library("lmerTest")
library("grid")
library("patchwork")
library("gridExtra")
library("sf")
library("rnaturalearth")
library("leaflet")
library("plotly")
library("leaflet.extras")
library("htmltools")
library("htmlwidgets")
library("ggmap")
library("osmdata")
library("tmap")
library("ggspatial")
```



```{r fonts, include=FALSE, message=FALSE, eval=FALSE, message=FALSE}
# Note that GhostScript needs to be installed at the system level for the PS files to be generated.
# MacOS users can use `brew install ghostscript`
# 
# Windows users can follow these directions:
# 1.	Go to the GhostScript website (https://www.ghostscript.com/download/gsdnld.html)
# 2.	Download the windows installer suitable for your machine
# 3.	Run the installer file which you downloaded and follow the prompts
# 4.	After running the installer click the windows "Start" button and type "Edit environment variables for your account" and open
# 5.	In the tab 'Advanced' click the button at the bottom 'Environment Variables...'
# 6.	Under 'System variables' find the variable 'Path', select 'Path' and click the 'Edit' button
# 7. 	Select a new line and copy the Ghostscript 'bin' folder location into the field.
# 7.1	If you installed ghostscript to the default folder location; then the folder location will likely be "C:\Program Files\gs\gs9.52\bin", the version number (9.52) may differ.
# 8.	Save and exit the environmental variables window
# This chunk is then run only if knitting on new computer that the files have not been generated on
# this is necessary to embed fonts in .eps files 
library("extrafont")
if (.Platform$OS.type == "windows") {
   font_import(pattern = "arial", prompt = FALSE)
   loadfonts(device = "postscript", quiet = TRUE)
} else {
   font_import(pattern = "Arial", prompt = FALSE)
   loadfonts(device = "postscript", quiet = TRUE)
}

extrafont::font_import()
```


## Set seed for reproducibility

```{r}
set.seed(42)
```


## Import cotton pathogenecity trial data

```{r}
dat_cotton <- read_excel("pathogenicity_trial_dat_fusarium.xlsx", sheet = 1) %>%
  mutate(crop = "cotton") %>%
  mutate(block = as.factor(block)) %>%
  mutate(root_severity_cotton = as.integer(root_severity_cotton)) %>%
  mutate(
    range = case_when(
      crop == "cotton" & root_severity_cotton == 0 ~ "0",
      crop == "cotton" & root_severity_cotton == 1 ~ "1–9",
      crop == "cotton" & root_severity_cotton == 2 ~ "10–30",
      crop == "cotton" & root_severity_cotton == 3 ~ "31–50",
        crop == "cotton" & root_severity_cotton == 4 ~ "51–80",
        crop == "cotton" & root_severity_cotton == 5 ~ "81–100",
      TRUE ~ NA_character_
    ),
    severity_midpoint = case_when(
        crop == "cotton" & root_severity_cotton == 0 ~ 0,
        crop == "cotton" & root_severity_cotton == 1 ~ 5,     # midpoint of 1–9
        crop == "cotton" & root_severity_cotton == 2 ~ 20,    # midpoint of 10–30
        crop == "cotton" & root_severity_cotton == 3 ~ 40.5,  # midpoint of 31–50
        crop == "cotton" & root_severity_cotton == 4 ~ 65.5,  # midpoint of 51–80
        crop == "cotton" & root_severity_cotton == 5 ~ 90.5,  # midpoint of 81–100
      TRUE ~ NA_real_
    ),
    species = case_when(
        crop == "cotton" & isolate_code == "F1" ~ "F. incarnatum",
        crop == "cotton" & isolate_code == "F16" ~ "F. proliferatum",
        crop == "cotton" & isolate_code == "F27" ~ "F. clavum",
        crop == "cotton" & isolate_code == "F30" ~ "F. solani",
        crop == "cotton" & isolate_code == "F39" ~ "F. flocciferum",
        crop == "cotton" & isolate_code == "F40" ~ "F. equiseti",
        crop == "cotton" & isolate_code == "F42" ~ "F. verticillioides",
      TRUE ~ "Control"
    ),
    severity_midpoint2 = severity_midpoint / 100, # convert to proportion
    severity_midpoint3 = ifelse(severity_midpoint2 == 0, 0.001, severity_midpoint2)  # change 0 to 0.001 to make it suitable for beta regression
  ) %>%
  rename(root_severity = root_severity_cotton) %>%
   relocate(species, .after = isolate_code) %>%
    mutate_if(is.character, as.factor) 
```


## Import corn pathogenicty data

```{r}
dat_corn <- read_excel("pathogenicity_trial_dat_fusarium.xlsx", sheet = 2) %>%
  mutate(crop = "corn") %>%
  mutate(block = as.factor(block)) %>%
  mutate(root_severity_corn = as.integer(root_severity_corn)) %>%
  mutate(
      range = case_when(
      crop == "corn" & root_severity_corn == 1 ~ "0",
      crop == "corn" & root_severity_corn == 2 ~ "1–9",
      crop == "corn" & root_severity_corn == 3 ~ "10–20",
      crop == "corn" & root_severity_corn == 4 ~ "21–40",
      TRUE ~ NA_character_
    ),
      severity_midpoint = case_when(
       crop == "corn" & root_severity_corn == 1 ~ 0,
       crop == "corn" & root_severity_corn == 2 ~ 5,     # midpoint of 1–9
       crop == "corn" & root_severity_corn == 3 ~ 15,    # midpoint of 10–20
       crop == "corn" & root_severity_corn == 4 ~ 30.5,  # midpoint of 20–40
      TRUE ~ NA_real_
    ),
    species = case_when(
       crop == "corn" & isolate_code == "F1" ~ "F. incarnatum",
       crop == "corn" & isolate_code == "F16" ~ "F. proliferatum",
       crop == "corn" & isolate_code == "F27" ~ "F. clavum",
       crop == "corn" & isolate_code == "F30" ~ "F. solani",
       crop == "corn" & isolate_code == "F39" ~ "F. flocciferum",
       crop == "corn" & isolate_code == "F40" ~ "F. equiseti",
       crop == "corn" & isolate_code == "F42" ~ "F. verticillioides",
      TRUE ~ "Control"
    ),
    severity_midpoint2 = severity_midpoint / 100, # convert to proportion
    severity_midpoint3 = ifelse(severity_midpoint2 == 0, 0.001, severity_midpoint2)  # change 0 to 0.001 to make it suitable for beta regression
  ) %>%
   relocate(species, .after = isolate_code) %>%
  rename(root_severity = root_severity_corn) %>%
    mutate_if(is.character, as.factor)
```


## Combine corn & cotton pathogenecity data

```{r}
dat_all <- bind_rows(dat_cotton, dat_corn) 

# Table for the ms
dat_table <- dat_all %>%
  group_by(crop, species) %>%
  summarise(mean_severity = mean(severity_midpoint2))

#openxlsx::write.xlsx(dat_table, "ms-table-root_severity.xlsx")
```


## Visualise cotton & corn pathogenicity 


```{r}
# Summarize by species and crop, converting to proportion
summary_data <- dat_all %>%
  group_by(species, crop) %>%
  summarise(
    mean_severity = mean(severity_midpoint, na.rm = TRUE) / 100,
    se = sd(severity_midpoint, na.rm = TRUE) / sqrt(n()) / 100
  )

# Plot
pathogenicity_plot <- ggplot(summary_data, aes(x = species, y = mean_severity, color = crop, group = crop)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(
    aes(ymin = mean_severity - se, ymax = mean_severity + se),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  labs(
    x = expression(italic("Fusarium") ~ "species"),
    y = "Mean disease severity (proportion)",
    color = "Crop"
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.20, 0.40, 0.60, 0.80, 1.0),
    labels = scales::number_format(accuracy = 0.01)
  ) +
  theme_pubclean(base_family = "Arial", base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  scale_color_brewer(palette = "Set1")

pathogenicity_plot
```


```{r}
ggsave(
  here("man", "figures/pathogenicity_plot.eps"),
  plot = pathogenicity_plot,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600,
  dev.off()
)


ggsave(
  here("man", "figures/pathogenicity_plot.png"),
  plot = pathogenicity_plot,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600
)
```


### Bar chart

```{r}
# Reorder species by total mean_severity
summary_data2 <- summary_data %>%
  group_by(species) %>%
  mutate(total_mean_severity = sum(mean_severity)) %>%
  ungroup() %>%
  mutate(species = fct_reorder(species, total_mean_severity))

# Now plot with reordered species
stacked_bar_plot <- ggplot(summary_data2, aes(x = species, y = mean_severity, fill = crop)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = expression(italic("Fusarium") ~ "species"),
    y = "Mean midpoint disease everity (proportion)",
    fill = "Crop"
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.20, 0.40, 0.60, 0.80, 1.0),
    labels = scales::number_format(accuracy = 0.01)
  ) +
  theme_pubclean(base_family = "Arial", base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  scale_fill_viridis_d(option = "cividis")

stacked_bar_plot

```


```{r}
ggsave(
    here("man", "figures/stacked_bar_plot.png"),
    plot = stacked_bar_plot,
    width = 6,
    height = 6,
    units = "in",
    dpi = 600
)
```


### Donut-chart

```{r}
dat_pie <- read_excel("survey_dat.xlsx", sheet = 1) %>%
  mutate(species = as.factor(species), count = as.integer(count))


# Compute position and labels
dat_pie <- dat_pie %>%
  arrange(desc(species)) %>%
  mutate(
    ypos = cumsum(percent) - 0.5 * percent,
    angle = 90 - (ypos / 100) * 360,
    hjust = ifelse(angle < -90, 1, 0),
    angle = ifelse(angle < -90, angle + 180, angle),
    label = paste0(species, " (", round(percent), "%)")
  )

# Donut chart using RColorBrewer Set3 palette
donut_chart <- ggplot(dat_pie, aes(x = 2, y = percent, fill = species)) +
  geom_bar(stat = "identity",
           width = 1,
           color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.6) +
  scale_fill_brewer(palette = "Paired") + 
  theme_pubclean(base_family = "Arial", base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  geom_text(aes(
    x = 1.6,
    y = ypos,
    label = label,
    angle = angle,
    hjust = hjust
  ),
  size = 3.2) 
  
  donut_chart
```


```{r}
ggsave(
  here("man", "figures/donut_chart.eps"),
  plot =   donut_chart,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600
)

ggsave(
  here("man", "figures/donut_chart.png"),
  plot =   donut_chart,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600
)
```



### Combine bar plot & donut chart


### Cobmine aerial pathogenecity plots

```{r}
Figure_2 <- donut_chart + stacked_bar_plot +
    plot_layout(ncol = 2) +
    plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(
        size = 15,
        face = "bold",
        family = "Arial"
    ))

Figure_2 
```


```{r}
ggsave(
    here("man", "figures/Figure_2.png"),
    plot = Figure_2,
    width = 10,
    height = 7,
    units = "in",
    dpi = 600
)

ggsave(
  here("man", "figures/Figure_2.eps"),
  plot = Figure_2,
  width = 10,
  height = 7,
  units = "in",
  dpi = 600,
device = cairo_ps
)

```



### Import lat-long information of the fields sampled

```{r}
# Import survey data
survey_dat <- read_excel("survey_dat.xlsx", sheet = 3) %>%
    mutate(
    lat = as.numeric(sapply(strsplit(as.character(lat_long), ","), `[`, 1)),
    long = as.numeric(sapply(strsplit(as.character(lat_long), ","), `[`, 2))
  ) %>%
  mutate(across(where(is.character), as.factor)) %>%
  group_by(county, sampling_date) %>%
  mutate(fields_sampled = n_distinct(county_code)) %>%
  ungroup()
```


## Add polygons

```{r}
# Read counties shape file
counties_dat <- st_read("c_05mr24.shp") %>%
  filter(STATE=="TX") %>%
  select(-c(LON, LAT))

# List counties sampled in our project
texas_counties <- c("Randall", "Hartley", "Hansford", "Moore", "Sherman", 
                    "Potter", "Hutchinson", "Dallam", "Deaf Smith", "Carson")



# Filter the shapefile for the sampled counties
texas_shapefile <- counties_dat %>% 
  filter(COUNTYNAME %in% texas_counties)


# Merge your survey data with the shapefile data for visualisation
merged_data <- left_join(texas_shapefile, survey_dat, by = c("COUNTYNAME" = "county"))

# Reproject the spatial data from NAD83 to WGS84 for visualisation
merged_data <- st_transform(merged_data, crs = 4326)

```

### Sampled counties in Texas High Plains

```{r}
showtext_auto()

# Read and project to Texas Albers Equal Area
counties_dat <- st_read("c_05mr24.shp") %>%
  filter(STATE == "TX") %>%
  st_transform(crs = 3083)  # NAD83 / Texas Centric Albers Equal Area

# Northern High Plains counties
northern_high_plains <- c(
  "Armstrong", "Briscoe", "Carson", "Collingsworth", "Dallam", "Deaf Smith",
  "Donley", "Gray", "Hall", "Hansford", "Hartley", "Hemphill", "Hutchinson",
  "Lipscomb", "Moore", "Ochiltree", "Oldham", "Potter", "Randall", "Roberts",
  "Sherman", "Wheeler"
)

# Sampled counties
sampled_counties <- c("Randall", "Hartley", "Hansford", "Moore", "Sherman", 
                      "Potter", "Hutchinson", "Dallam", "Deaf Smith", "Carson")

# Filter and annotate sampled counties
north_hp_map <- counties_dat %>%
  filter(COUNTYNAME %in% northern_high_plains) %>%
  mutate(sampled = ifelse(COUNTYNAME %in% sampled_counties, "Sampled", "Not Sampled")) %>%
    mutate_if(is.character, as.factor)

# Get label points for counties
label_points <- st_point_on_surface(north_hp_map)

# Create map
sampling_area_plot <- ggplot(data = north_hp_map) +
    geom_sf(aes(fill = sampled), color = "black", size = 0.2) +
    geom_sf_text(data = label_points, aes(label = COUNTYNAME), size = 2.5) +
    scale_fill_manual(
        values = c("Sampled" = "#0072B2", "Not Sampled" = "#D55E00"),
        labels = c("Sampled", "Not Sampled"),
        name = "Counties"
    ) +
    annotation_scale(location = "bl", width_hint = 0.3, text_cex = 0.7, line_width = 0.3) +
    annotation_north_arrow(
        location = "tr", which_north = "true",
        pad_x = unit(0, "cm"), pad_y = unit(0, "cm"),
        style = north_arrow_fancy_orienteering
    ) +
    coord_sf(
        crs = st_crs(3083),
        datum = st_crs(4326),
        label_graticule = "SW",
        xlim = c(st_bbox(north_hp_map)[["xmin"]], st_bbox(north_hp_map)[["xmax"]]),
        ylim = c(st_bbox(north_hp_map)[["ymin"]], st_bbox(north_hp_map)[["ymax"]])
    ) +
    theme_pubclean(base_size = 15) +
    theme(
        legend.position = "bottom",
        axis.title = element_blank(),
        axis.text = element_text(size = 8),
        axis.ticks = element_line(size = 0.3)
        # panel.grid.major = element_line(color = "gray80", size = 0.2)
    )

sampling_area_plot

```




```{r}
ggsave(
  here("man", "figures/sampling_area_plot.pdf"),
  plot = sampling_area_plot,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600
)

ggsave(
  here("man", "figures/sampling_area_plot.png"),
  plot = sampling_area_plot,
  width = 8,
  height = 8,
  units = "in",
  dpi = 600
)
```
