---
title: "Analysis"
author: "Ahsan"
date: "2023-06-26"
output: html_document
---

```{r setup, include=FALSE}
library("grDevices")
library("tidyverse")
library("viridis")
library("showtext")
library("here")
library("ggpubr")
library("extrafont")
library("kableExtra")
library("tidyr")
library("lubridate")
library("readxl")
library("ggstatsplot")
library("pairwiseComparisons")
library("multcompView")
library("ggthemes")
library("emmeans")
library("multcomp")
library("car")
library("DHARMa")
library("lmerTest")
library("grid")
library("patchwork")
library("gridExtra")
library("agricolae")
```


## Set seed for reproducibility

```{r}
set.seed(42)
```


# Cotton

## Import data

```{r}
dat <- read_excel("pathogenicity_trial_dat_fusarium.xlsx", sheet = 1) %>%
  mutate(block = as.factor(block)) %>%
  mutate(root_severity_cotton = as.integer(root_severity_cotton)) %>%
  mutate(
    range = case_when(
      root_severity_cotton == 0 ~ "0",
      root_severity_cotton == 1 ~ "1–9",
      root_severity_cotton == 2 ~ "10–30",
      root_severity_cotton == 3 ~ "31–50",
      root_severity_cotton == 4 ~ "51–80",
      root_severity_cotton == 5 ~ "81–100",
      TRUE ~ NA_character_
    ),
    severity_midpoint = case_when(
      root_severity_cotton == 0 ~ 0,
      root_severity_cotton == 1 ~ 5,     # midpoint of 1–9
      root_severity_cotton == 2 ~ 20,    # midpoint of 10–30
      root_severity_cotton == 3 ~ 40.5,  # midpoint of 31–50
      root_severity_cotton == 4 ~ 65.5,  # midpoint of 51–80
      root_severity_cotton == 5 ~ 90.5,  # midpoint of 81–100
      TRUE ~ NA_real_
    ),
    species = case_when(
      isolate_code == "F1" ~ "F. incarnatum",
      isolate_code == "F16" ~ "F. proliferatum",
      isolate_code == "F27" ~ "F. clavum",
      isolate_code == "F30" ~ "F. solani",
      isolate_code == "F39" ~ "F. flocciferum",
      isolate_code == "F40" ~ "F. equiseti",
      isolate_code == "F42" ~ "F. verticillioides",
      TRUE ~ "Control"
    ),
    severity_midpoint2 = severity_midpoint / 100, # convert to proportion
    severity_midpoint3 = ifelse(severity_midpoint2 == 0, 0.001, severity_midpoint2)  # change 0 to 0.001 to make it suitable for beta regression
  ) %>%
   relocate(species, .after = isolate_code) %>%
    mutate_if(is.character, as.factor)

  # dplyr::group_by(assessment, treatment, block) %>%
  # summarise(mean_severity_trt = mean(severity_mp),
  #           sd_severity_trt = sd(severity_mp))
```


## Visualize data

```{r}
cotton_plot <- ggplot(dat, aes(x = fct_reorder(species, severity_midpoint, .fun =mean), y = severity_midpoint)) +
  geom_boxplot(outlier.shape = NA, fill = "#91B493", alpha = 0.7, color = "black") +
  geom_jitter( width = 0.2, size = 2, show.legend = FALSE) +
  labs(
    x = expression(italic("Fusarium") ~ "species"),
    y = "Disease severity (midpoint %)"
  ) +
  scale_y_continuous(breaks = seq(0, 100, by = 25)) +
  expand_limits(y = c(0, 100)) +
  theme_pubclean(base_family = "Arial", base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold") 
  ) 
  
cotton_plot
# "#bc3032"
```


## Run a linear model 

```{r}
m1 <- aov(severity_midpoint ~ species, data= dat)

summary(m1)
```


## Check normality and homocedasticity tests

```{r}
plot(simulateResiduals(m1))
```

Since both normality and homoscedasticity assumptions are violated in our initial model, we will halt further analysis using this model. Instead, we will explore an alternative approach that includes data transformation to proportion data and then logit transformation. 

## Apply transformation (proportion & logit)

```{r, message=FALSE}
m2 <- lmer(logit(severity_midpoint3) ~ species + (1 | block), data = dat)

summary(m2)
```


## Check for normality and homocedasticity tests

```{r}
plot(simulateResiduals(m2))
```

Model is OK.

## Plot model

```{r}
# Means comparison on the response (back-transformed scale)
means_m2 <- emmeans(m2, "species", type = "response")

# Compact Letter Display
cld_df_cotton <- cld(means_m2, Letters = letters)

# Prepare for plotting
means_df_cotton <- as.data.frame(means_m2) %>%
    arrange(desc(response))

# Make sure species levels match for consistent ordering
cld_df_cotton$species <- factor(cld_df_cotton$species, levels = levels(means_df_cotton$species))

# Merge CLD letters with emmeans results
plot_df_cotton <- left_join(means_df_cotton, cld_df_cotton[, c("species", ".group")], by = "species")

# Display p values
pairs(means_m2)

# Clean up group letters: remove whitespace
plot_df_cotton <- plot_df_cotton %>%
  mutate(.group = gsub(" ", "", .group), # remove spaces
         label_y = upper.CL + 0.03)        # position letters just above error bar


cotton_model_plot <- ggplot(plot_df_cotton, aes(x = reorder(species, response), y = response)) +
  geom_point(size = 3, color = "#0072B2") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.15, color = "#0072B2") +
  geom_text(aes(y = label_y, label = .group), size = 5, color = "black") +
  labs(
    x = expression(italic("Fusarium") ~ "species"),
    y = "Disease severity (proportion)"
  ) +
  theme_pubclean(base_family = "Arial", base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   ylim(0, 1)

cotton_model_plot

# # Fixed effect summary
# coefs <- summary(m2)$coefficients %>%
#   as.data.frame() %>%
#   mutate(term = rownames(.)) %>%
#   #filter(term != "(Intercept)") %>%
#   mutate(
#     species = gsub("species", "", term),
#     logit_mean = Estimate + summary(m2)$coefficients["(Intercept)", "Estimate"],
#     lower.CL = logit_mean - 1.96 * `Std. Error`,
#     upper.CL = logit_mean + 1.96 * `Std. Error`,
#     sig = case_when(
#       `Pr(>|t|)` < 0.05 ~ "*",
#       `Pr(>|t|)` < 0.1 ~ ".",
#       TRUE ~ ""
#     )
#   )
# # rename intercept to control 
# coefs$species <- fct_recode(coefs$species, "Control" = "(Intercept)")
# 
# # Convert to probability scale
# logit_to_prob <- function(x) exp(x) / (1 + exp(x))
# 
# coefs <- coefs %>%
#   mutate(
#     prob_mean = logit_to_prob(logit_mean),
#     prob_lower = logit_to_prob(lower.CL),
#     prob_upper = logit_to_prob(upper.CL)
#   )
# 
# # Assign group letters (you can automate this too, but keeping your manual logic here)
# coefs$group <- "a"
# coefs$group[coefs$species == "F. proliferatum"] <- "b"
# coefs$group[coefs$species  == "F. flocciferum"] <- "b"
# 
# # Add a y-position just above the upper confidence interval
# coefs <- coefs %>%
#   mutate(label_y = prob_upper + 0.03)  
# 
# 
# Plot on probability scale
# cotton_model_plot <- ggplot(coefs, aes(x = reorder(species, prob_mean), y = prob_mean)) +
#   geom_point(size = 3, color = "#0072B2") +
#   geom_errorbar(aes(ymin = prob_lower, ymax = prob_upper), width = 0.15, color = "#0072B2") +
#   geom_text(aes(y = label_y, label = group), size = 6, fontface = "bold") +
#   labs(
#     x = expression(italic("Fusarium") ~ "species"),
#     y = "Disease severity (proportion)"
#   ) +
#   theme_pubclean(base_family = "Arial", base_size = 12) +
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#   ) +
#   ylim(0, 1)
# 
# cotton_model_plot

```


# Corn

## Import data

```{r}
dat_corn <- read_excel("pathogenicity_trial_dat_fusarium.xlsx", sheet = 2) %>%
  mutate(block = as.factor(block)) %>%
  mutate(root_severity_corn = as.integer(root_severity_corn)) %>%
  mutate(
    corn_range = case_when(
      root_severity_corn == 1 ~ "0",
      root_severity_corn == 2 ~ "1–9",
      root_severity_corn == 3 ~ "10–20",
      root_severity_corn == 4 ~ "21–40",
      TRUE ~ NA_character_
    ),
    corn_severity_midpoint = case_when(
      root_severity_corn == 1 ~ 0,
      root_severity_corn == 2 ~ 5,     # midpoint of 1–9
      root_severity_corn == 3 ~ 15,    # midpoint of 10–20
      root_severity_corn == 4 ~ 30.5,  # midpoint of 20–40
      TRUE ~ NA_real_
    ),
    species = case_when(
      isolate_code == "F1" ~ "F. incarnatum",
      isolate_code == "F16" ~ "F. proliferatum",
      isolate_code == "F27" ~ "F. clavum",
      isolate_code == "F30" ~ "F. solani",
      isolate_code == "F39" ~ "F. flocciferum",
      isolate_code == "F40" ~ "F. equiseti",
      isolate_code == "F42" ~ "F. verticillioides",
      TRUE ~ "Control"
    ),
    corn_severity_midpoint2 = corn_severity_midpoint / 100, # convert to proportion
    corn_severity_midpoint3 = ifelse(corn_severity_midpoint2 == 0, 0.001, corn_severity_midpoint2)  # change 0 to 0.001 to make it suitable for beta regression
  ) %>%
   relocate(species, .after = isolate_code) %>%
    mutate_if(is.character, as.factor)
```


## Visualize corn pathogenicity data

```{r}
corn_plot <- ggplot(dat_corn, aes(x = fct_reorder(species, corn_severity_midpoint, .fun =mean), y = corn_severity_midpoint)) +
  geom_boxplot(outlier.shape = NA, fill = "#91B493", alpha = 0.7, color = "black") +
  geom_jitter( width = 0.2, size = 2, show.legend = FALSE) +
  labs(
    x = expression(italic("Fusarium") ~ "species"),
    y = "Disease severity (midpoint %)"
  ) +
  scale_y_continuous(breaks = seq(0, 100, by = 25)) +
  expand_limits(y = c(0, 100)) +
  theme_pubclean(base_family = "Arial", base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold") 
  ) 
  
corn_plot
# "#bc3032"
```


## Run a linear model 

```{r}
m1_corn <- aov(root_severity_corn ~ species, data= dat_corn)

summary(m1_corn)
```


## Check for normality and homocedasticity tests

```{r}
plot(simulateResiduals(m1_corn))
```

Diagnostics test suggest that a simple linear model can be used.


## Apply transformation (proportion & logit)

```{r}
m2_corn <- lmer(logit(corn_severity_midpoint3) ~ species + (1 | block), data = dat_corn)

summary(m2_corn)
```

Both models meet the assumptions, we will go with m2_corn to account for the random effect of block.

## Check for normality and homocedasticity tests

```{r}
plot(simulateResiduals(m2_corn))
```


## Plot model using emmeans R package 

Both assumptions have been met. We can proceed to use a means comparison test to determine which treatments differ from one another, with the assistance of the `emmeans` package. It’s important to note that the results are displayed in the original scale after transformation, specifically when using type = "response".

### Means comparison

```{r}
# Means comparison on the response (back-transformed scale)
means_m2_corn <- emmeans(m2_corn, "species", type = "response")

# Compact Letter Display
cld_df_corn <- cld(means_m2_corn, Letters = letters)

# Prepare for plotting
means_df_corn <- as.data.frame(means_m2_corn) %>%
  arrange(desc(response))

# Make sure species levels match for consistent ordering
cld_df_corn$species <- factor(cld_df_corn$species, levels = levels(means_df_corn$species))

# Merge CLD letters with emmeans results
plot_df_corn <- left_join(means_df_corn, cld_df_corn[, c("species", ".group")], by = "species")
```


### Plot corn pathogenicity trial

```{r}
# Clean up group letters: remove whitespace
plot_df_corn <- plot_df_corn %>%
  mutate(
    response = ifelse(species == "F. solani", response + 0.001, response), # added .001 to soani to ensure control is dispalyed at the left most side.
    .group = gsub(" ", "", .group),  # remove spaces
    label_y = upper.CL + 0.05        # position letters just above error bar
  )

# Plot
corn_model_plot <- ggplot(plot_df_corn, aes(x = reorder(species, response), y = response)) +
  geom_point(size = 3, color = "#0072B2") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.15, color = "#0072B2") +
  geom_text(aes(y = label_y, label = .group), size = 5, color = "black") +
  labs(
    x = expression(italic("Fusarium") ~ "species"),
    y = "Disease severity (proportion)"
  ) +
  theme_pubclean(base_family = "Arial", base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   ylim(0, 1)

corn_model_plot
```


## Combine cotton & corn model plots

```{r}
# Cotton plot
cotton_model_plot <- ggplot(plot_df_cotton, aes(x = reorder(species, response), y = response)) +
  geom_point(size = 3, color = "#E69F00") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.15, color = "#E69F00") +
  geom_text(aes(y = label_y, label = .group), size = 5, color = "black") +
  labs(
    x = "",
    y = ""
  ) +
  theme_pubclean(base_family = "Arial", base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   ylim(0, 1)


# Corn plot
corn_model_plot <- ggplot(plot_df_corn, aes(x = reorder(species, response), y = response)) +
  geom_point(size = 3, color = "#E69F00") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.15, color = "#E69F00") +
  geom_text(aes(y = label_y, label = .group), size = 5, color = "black") +
  labs(
    x = "",
    y = ""
  ) +
  theme_pubclean(base_family = "Arial", base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   ylim(0, 1)

# Add annotations to each individual plot
cotton_model_plot <- cotton_model_plot + 
  plot_annotation(tag_levels = 'A') +
  theme(plot.tag = element_text(size = 15, face = "bold", family = "Arial"))

corn_model_plot <- corn_model_plot + 
  plot_annotation(tag_levels = 'B') +
  theme(plot.tag = element_text(size = 15, face = "bold", family = "Arial"))

# Shared x-axis label (text)
shared_x <- textGrob(
  label = expression(italic("Fusarium") ~ "species"),
  gp = gpar(fontsize = 14, fontfamily = "Arial"),
  hjust = 0.5,
  x = 0.5
)

# Shared y-axis label (text)
shared_y <- textGrob(
  "Disease severity (proportion)",
  gp = gpar(fontsize = 14, fontfamily = "Arial"),
  rot = 90,
  hjust = 0.5,
  vjust = 0.5
)

# Combine the plots and shared axis labels
combined_modelling_plot <- cotton_model_plot + corn_model_plot +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(
    size = 15,
    face = "bold",
    family = "Arial"
  ))


# Convert the combined plot to a grob
combined_grob <- patchworkGrob(combined_modelling_plot)

# Use grid.arrange to add shared x and y labels
Figure_5 <- grid.arrange(
  combined_grob, 
  left = shared_y,  # Left y-axis label
  bottom = shared_x  # Bottom x-axis label
)

Figure_5
```



```{r}
ggsave(
  here("man", "figures/Figure_5.eps"),
  plot = Figure_5,
  width = 8,
  height = 6,
  units = "in",
  dpi = 600,
device = cairo_ps
)


png("Figure_5.png", width = 8, height = 6, units = "in", res = 600)
grid.arrange(
    combined_grob, 
    left = shared_y,
    bottom = shared_x
)
dev.off()
```


## Cotton aerial parts severity

### Import data

```{r}
dat_aerial <- read_excel("data/pathogenicity_trial_dat_fusarium.xlsx", sheet = 3) %>%
  mutate(block = as.factor(block)) %>%
  mutate(severity = as.integer(severity)) %>%
  mutate(species = case_when(
    isolate_code  == "F1" ~ "F. incarnatum",
    isolate_code  == "F16" ~ "F. proliferatum",
    isolate_code  == "F27" ~ "F. clavum",
    isolate_code  == "F30" ~ "F. solani",
    isolate_code  == "F39" ~ "F. flocciferum",
    isolate_code  == "F40" ~ "F. equiseti",
    isolate_code == "F42" ~ "F. verticillioides",
    TRUE ~ "Control"
  )) %>%
  relocate(species, .after = isolate_code) %>%
  mutate_if(is.character, as.factor)
```


## Calculate the area under the disase progress stairs (AUDPS)

```{r}
dat_aerial2 <- dat_aerial %>%
  mutate(julian_day = yday(assessment_date)) %>%
 group_by(species, julian_day) %>%
  summarise(mean_severity = mean(severity), .groups = "drop")

audps_dat <- dat_aerial2 %>%
  group_by(species) %>%
  summarise(AUDPS = audps(mean_severity, julian_day), .groups = "drop") %>%
  arrange(AUDPS)


#openxlsx::write.xlsx(audps_dat, "ms-table_audps.xlsx")
```


## Plot AUDPS

```{r}
# Create a named vector of expression labels to italicize species
species_labels <- c(
  "Control" = "Control",
  "F. verticillioides" = expression(italic("F. verticillioides")),
  "F. clavum" = expression(italic("F. clavum")),
  "F. flocciferum" = expression(italic("F. flocciferum")),
  "F. equiseti" = expression(italic("F. equiseti")),
  "F. proliferatum" = expression(italic("F. proliferatum")),
  "F. solani" = expression(italic("F. solani")),
  "F. incarnatum" = expression(italic("F. incarnatum"))
)

audps_plot <- ggplot(audps_dat, aes(x = reorder(species, AUDPS), y = AUDPS)) +
    geom_col(fill = "#1f78b4") +
    coord_flip() +
    scale_x_discrete(labels = species_labels) +  
    labs(x = expression(italic("Fusarium") ~ "species"), y = "Disease severity (AUDPS)") +
    theme_pubclean(base_family = "Arial", base_size = 14)
  
  audps_plot
```


## Analyze the area under the disease progress curve

```{r}
# Calculate AUDPS by block
audps_dat2 <- dat_aerial %>%
  mutate(julian_day = yday(assessment_date)) %>%
  group_by(species, block, julian_day) %>%
  summarise(mean_severity = mean(severity), .groups = "drop") %>%
  group_by(species, block) %>%
  summarise(AUDPS = audps(mean_severity, julian_day), .groups = "drop")

#openxlsx::write.xlsx(audps_dat2, "ms-table_audps2.xlsx")
```


## Fit model 

```{r}
m2_c_aerial <- lmer(AUDPS ~ species + (1 | block), data = audps_dat2)

summary(m2_c_aerial)
```

## Check for normality and homocedasticity tests

```{r}
plot(simulateResiduals(m2_c_aerial))
```


## Get estimated means and then plot model 

```{r}
# Means comparison on the response
means_m_c_aerial <- emmeans(m2_c_aerial, "species")

# Compact Letter Display
cld_df_m_c <- cld(means_m_c_aerial, Letters = letters)

# Prepare for plotting
means_df_m_c <- as.data.frame(means_m_c_aerial) %>%
  arrange(desc(emmean))

# Make sure species levels match for consistent ordering
cld_df_m_c$species <- factor(cld_df_m_c$species, levels = levels(means_df_m_c$species))


# Drop all columns from cld_df_m_c *except* species and .group
plot_df_m_c <- left_join(means_df_m_c, cld_df_m_c[, c("species", ".group")], by = "species")


# Clean up group letters: remove whitespace
plot_df_m_c2 <- plot_df_m_c %>%
  mutate(.group = gsub(" ", "", .group), # remove spaces
         label_y = upper.CL + 0.03)        # position letters just above error bar


mc_model_plot <- ggplot(plot_df_m_c2, aes(x = reorder(species, emmean), y = emmean)) +
    geom_point(size = 3, color = "#0072B2") +
    geom_errorbar(aes(
        ymin = pmax(lower.CL, 0),  # Prevent error bar from going below zero
        ymax = upper.CL
    ), width = 0.15, color = "#0072B2") +
    geom_text(aes(y = label_y + 2, label = .group), size = 5, color = "black") +  # Added space here
    labs(
        x = expression(italic("Fusarium") ~ "species"),
        y = "Disease severity (AUDPS)"
    ) +
    theme_pubclean(base_family = "Arial", base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    coord_cartesian(ylim = c(0, 45))

mc_model_plot
```

### Cobmine aerial pathogenecity plots

```{r}
Figure_4 <- audps_plot + mc_model_plot +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(
    size = 15,
    face = "bold",
    family = "Arial"
  ))

Figure_4
```



```{r}
ggsave(
  here("man", "figures/Figure_4.eps"),
  plot =  Figure_4,
  width = 9,
  height = 5,
  units = "in",
  dpi = 600,
  device = cairo_ps
)

ggsave(
  here("man", "figures/Figure_4.png"),
  plot =  Figure_4,
  width = 9,
  height = 4.5,
  units = "in",
  dpi = 600
)
```



